\texttt{davos}: a Python package ``smuggler'' for constructing lightweight reproducible notebooks

### 2022-10-11 paper/main.tex

\subsection{Software architecture}
The \texttt{davos} package is structured as two sub-packages: a set of ``core" modules that implement...

% Give a short overview of the overall software architecture; provide a pictorial component overview or similar (if possible). If necessary provide implementation details.

\subsection{Software functionalities}%\label{sec:functionalities}

\subsubsection{The \texttt{smuggle} statement}
Importing \texttt{davos}\comment{in a Jupyter notebook} enables an additional Python keyword: ``\texttt{smuggle}".
The \texttt{smuggle} statement can be used as a drop-in replacement for Python's built-in \texttt{import} statement to load libraries, modules, and other objects into the current namespace.
However, whereas \texttt{import} will fail if the requested package is not installed locally, \texttt{smuggle} statements can handle missing packages on the fly.
If a smuggled package does not exist in the local environment, \texttt{davos} will install it, expose its contents to Python's \texttt{import} machinery, and load it into the namespace for immediate use.

\subsubsection{The onion comment}
For greater control over the behavior of \texttt{smuggle} statements, \texttt{davos} defines an additional construct called the \textit{onion comment}. An onion comment is a special type of inline comment that may be placed on a line containing a \texttt{smuggle} statement to customize how \texttt{davos} searches for the smuggled package locally and, if necessary, how it should be installed. Onion comments follow a simple syntax based on the ``type comment" syntax introduced in PEP 484 \cite{vanREtal14} and are designed to make managing packages via \texttt{davos} intuitive and familiar. To construct an onion comment, simply provide the name of the installer program (e.g., \texttt{pip}) and the same arguments one would use to install the package as desired manually via the command line (see Fig.~\ref{fig:snippet1}).

\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{snippets/snippet1.pdf}
\caption{\small \textbf{FILL THIS IN...}}
\label{fig:snippet1}
\end{figure}

>>>>

\subsection{Software architecture}
The \texttt{davos} package consists of two interdependent subpackages. The first, \texttt{davos.core}, comprises a set of modules that implement\comment{provide?} the bulk of the package's core functionality, including pipelines for installing and validating packages, custom parsers for the \texttt{smuggle} statement (see Section \ref{subsec:smuggle}) and onion comment (see Section \ref{subsec:onion}), and a runtime interface for configuring \texttt{davos}'s behavior (see Section \ref{subsec:config}). However, certain critical\comment{other important} aspects of this functionality require (often substantially) different implementation approaches depending on various properties of the notebook environment in which \texttt{davos} is used (e.g., whether the frontend is provided by Jupyter or Google Colaboratory, or which version of IPython \cite{PereGran07} is used by the notebook kernel). To deal with this, environment-dependent parts of core features and behaviors are isolated and abstracted to ``helper functions'' in the \texttt{davos.implementations} subpackage. This second subpackage defines multiple, interchangeable versions of each helper function, organized into\comment{per-environment} modules by the conditions that trigger their use. At runtime, \texttt{davos} detects various features in the notebook environment and selectively imports a single version of each helper function into the top-level \texttt{davos.implementations} namespace, allowing \texttt{davos.core} modules to access the correct implementations for the current notebook environment from a single, constant location. An additional benefit of this design pattern is that it makes adding support for new or updated notebook variants to \texttt{davos} in the future relatively easy.

\comment{
- js_functions.py?
- how parser is registered and deregistered?
- stub files?
- test suite?
- packaged with new PEP \_\_\_ standard (pyproject.toml + setup.cfg; no setup.py)?
}
% Give a short overview of the overall software architecture; provide a pictorial component overview or similar (if possible). If necessary provide implementation details.

\subsection{Software functionalities}

\subsubsection{The \texttt{smuggle} statement}\label{subsec:smuggle}
Importing \texttt{davos} in a Jupyter notebook enables an additional Python keyword: ``\texttt{smuggle}''.
The \texttt{smuggle} statement can be used as a drop-in replacement for Python's built-in \texttt{import} statement to load libraries, modules, and other objects into the current namespace.
However, whereas \texttt{import} will fail if the requested package is not installed locally, \texttt{smuggle} statements can handle missing packages on the fly.
If a smuggled package does not exist in the local environment, \texttt{davos} will install it, expose its contents to Python's \texttt{import} machinery, and load it into the namespace for immediate use.

\subsubsection{The onion comment}\label{subsec:onion}
For greater control over the behavior of \texttt{smuggle} statements, \texttt{davos} defines an additional construct called the ``onion comment''. An onion comment is a special type of inline comment that may be placed on a line containing a \texttt{smuggle} statement to customize how \texttt{davos} searches for the smuggled package locally and, if necessary, downloads and installs it. Onion comments follow a simple syntax based on the ``type comment'' syntax introduced in PEP 484 \cite{vanREtal14}, and are designed to make managing packages with \texttt{davos} intuitive and familiar. To construct an onion comment, simply provide the name of the installer program (e.g., \texttt{pip}) and the same arguments one would use to manually install the package as desired via the command line\comment{(see Fig.~\ref{fig:snippets})} (see Fig.~\ref{fig:snippets}, lines 3--10 for examples).
%For examples of \texttt{smuggle} statements and accompanying onion comments, see Figure \ref{fig:snippets}.
%Figure \ref{fig:snippets} provides a number of examples of \texttt{smuggle} statements accompanied by

Onion comments are useful when smuggling a package whose distribution name (i.e., the name used when installing it) is different from its top-level module name (i.e., the name used when importing it; e.g., Fig.~\ref{fig:snippets}, lines 12--13). However, the most powerful use of the onion comment is making \texttt{smuggle} statements \textit{version-sensitive}. If an onion comment includes a version specifier \cite{CoghStuf13} (e.g., Fig.~\ref{fig:snippets}, lines 15--19), \texttt{davos} will ensure that the version of the package loaded into the notebook matches the specific version requested, or satisfies the given version constraints. If the smuggled package exists locally, \texttt{davos} will extract its version info from its metadata and compare it to the specifier provided. If the two are incompatible (or no local installation is found), \texttt{davos} will install and load a suitable version of the package instead. Onion comments can similarly be used to smuggle specific VCS references (e.g., Git \cite{TorvHama05}  branches, commits, tags, etc.; Fig.~\ref{fig:snippets}, lines 21--22).

%If an onion comment includes a version specifier (\cite{CoghStuf13}; e.g., Fig.~\ref{fig:snippets}, lines \_\_\_-\_\_\_), \texttt{davos} will first check whether the smuggled package exists locally, and if so, extract its version info from its metadata to compare against the requested version. If the installed version doesn't satisfy the given constraint (or if no local version is found), \texttt{davos} will install and load a suitable version instead. This guarantees the notebook is always run with a specific (or constrained) version of a particular package. Similarly, \texttt{davos}
%If the package exists locally, \texttt{davos} will extract its version from metadata before loading it to determine its version, and if it does not match the requested version (or if no local installation is found),
%If an onion comment includes a version specifier \cite{CoghStuf13}, \texttt{davos} will check whether the smuggled package is installed locally (as usual), and if so, additionally parse its metadata to determine
%Adding a version specifier \cite{CoghStuf13} to an onion comment guarantees the version loaded into the notebook will matches the version specified

\texttt{davos} processes onion comments internally before forwarding arguments to the installer program. In addition to preventing onion comments from being used as a vehicle for shell injection attacks, this allows \texttt{davos} take certain logical actions when particular arguments are passed (e.g., Fig.~\ref{fig:snippets}, lines 24--28). For example, the \texttt{--force-reinstall}, \texttt{-I}/\texttt{--ignore-installed}, and \texttt{-U}/\texttt{--upgrade} flag will all cause \texttt{davos} to skip searching for a smuggled package locally before installing a new copy; \texttt{--no-input} will disable \texttt{davos}'s input prompts in addition to the installer program's; and installing a package into \texttt{<dir>} with \texttt{--target <dir>} will cause \texttt{dir} to be prepended to the module search path (\texttt{sys.path}), if necessary, so the package can be imported.

\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{figs/snippets.pdf}
\caption{\small Example \texttt{smuggle} statements and accompanying onion comments.}
\label{fig:snippets}
\end{figure}
